\chapter{Developer documentation}
\label{ch:impl}

\section{The task}

My task was to create a generic integration between Wrangler and trending IDEs (such as Visual Studio Code) by implementing the Language Server Protocol (LSP) in the Wrangler project.

The \emph{Wrangler Language Server} is a new component that provides the functionalities of the Wrangler refactoring tool (such as function and module renaming, function extraction and elimination) to any LSP-compatible code editor. We choose to do this by extending the \emph{Erlang Language Server}, an existing language server implementing the protocol.

In particular, \emph{Wrangler LS} is part of the Wrangler project, thus it is written in Erlang, and it is started and run by the \emph{Erlang LS}.

For this task, I had to partly redesign Wrangler’s API, and tailor the implementation of refactoring interactions to take full advantage of the Language Server Protocol’s capabilities. 

This work involved the investigation of how non-standard inputs and interactions can be implemented on top of the basic features of the protocol.

It was an additional requirement that the Wrangler LS had to be idiomatic, and it had to be conventional with the Erlang Language Server's structure. 

Furthermore, the solution is expected to require no manual setup apart from doing some simple configuration.

\section{Design choices}

\subsection{Language Server Protocol}

The Language Server Protocol seemed to be a great choice to give support for the modern development tools, however, LSP's support for refactorings is stuck in its early stage.

One cause is that the refactorings are very dependent on the programming languages and the other is that LSP do not want to limit the user interfaces how text editors handle user inputs.

Thus, LSP not only does not offer a protocol for the different refactorings, it does not even give a possibility to request various inputs from the user.

We had to come up with a way to acquire every data Wrangler refactorings need. Besides the ones that can be calculated from the cursor position or gained from the configuration, some refactorings usually need a new atom, variable name or file name to operate, and some need even more complex and interdependent input/output methods. 

In the former case, we could collect user input through a basic text field handled by the Visual Studio Code extension's middleware.
This is defined outside of the language server, using the text editor's extension APIs, in typescript.
It is called a middleware because when the user starts a refactoring, while the command is being sent from the client to the server, it catches the commands of these refactorings and modify their arguments based on the user input.
The drawback for this is that Wrangler has to come up with similar solutions for every code editor. If one of the editors does not offer any similar way to add user input to the command arguments, default values are used (for example "\tt NewVar\rm " as a new variable name).  

In the latter case, the procedure we had to give an option for the users was the following:
\begin{enumerate}
    \item the user choose the refactoring just like any other refactoring;
    \item highlight every code snippet where the refactoring can be executed;
    \item give an option for the user to apply the changes at each of the highlighted code snippets.
\end{enumerate}

This can be done with some kind of form that enables us to show the refactoring candidates along with some kind of buttons. We call the solution expressed below Wrangler Forms.

With LSP features, namely highlight and semantic tokens we are able to differentiate the refactoring candidates from the rest of the code. Highlight changes the text's background color in a given range, whereas semantic tokens changes the foreground color.

\begin{note}
    As its name reflects, semantic tokens meant to be used to inform the user about specific semantic information with color coding.
    Because Erlang LS have not implemented this feature yet, this does not cause any problem, but it will be the Erlang LS's responsibility to give preference to Wrangler's tokens when Wrangler Forms are used. Highlights have the same behaviour now.
\end{note}

With code lenses, clickable text pieces can be appended to the text file at the given places. We can use these to add button-like elements next to each refactoring candidate. This way, the user can select all the candidates he/she wants to refactor, or exit the form.

The drawback we had to compromise on was that manually changing the file would break the form. To avoid this, we put warnings for the users.

This method is good because, unlike the previous solution for text inputs, the whole process is under the Language Server Protocol, therefore it can be used in any text editor.

\subsection{The Erlang Language Server}

\todo{Why extending ELS, following similar structure}

\section{Used language features defined in the Language Server Protocol}

Apart from the base protocol for the communication between language servers and development tools, the Language Server Protocol defines various language features. These include requests like \textit{go to implementation, completion, and formatting}. To use some of these features, both the language servers and code editors need to support these one by one. Registering these are defined also in the protocol.\cite{LSP}

In the \emph{Wrangler Language Server}, the following LSP features are used:

\begin{itemize}
    \item textDocument/codeLens
    \item textDocument/codeAction
    \item textDocument/documentHighlight
    \item textDocument/semanticTokens/full
    \item workspace/executeCommand
    \item window/showMessage
    \todo{What do we use these for?}
\end{itemize}

These are well defined in the protocol \cite{LSP}, we only discuss their usage later in the paper.

\begin{note}
The development tool presented in this paper, Visual Studio Code supports all the above LSP features. 
\end{note}

The protocol version \emph{3.16} were used during my thesis.

\section{Implementation}

The current implementation of the \emph{Wrangler Language Server} consists of three parts:

\begin{itemize}
    \item Extending the Erlang Language Server to supplement their LSP responses with the Wrangler Language Server's features.
    \item Creating a new component in Wrangler to provide additional functionalities for Erlang LS. This is the Wrangler LS itself.
    \item Adding a middleware in the Erlang Language Server's Visual Studio Code extension to handle additional user inputs. 
\end{itemize}

\todo{diagram??}

\subsection{Extending the Erlang Language Server}

The Erlang Language Server introduced separate erlang modules called \tt provider\rm s for each language feature (for example, code lenses). These providers implement the \tt els\_provider\rm\ erlang behaviour and handle the corresponding requests coming from the client. \todo{diagram??}

In the case of code lenses, the different lenses are placed in their own modules. Each code lens module implements the \tt els\_code\_lens\rm\  erlang behaviour. The provider gathers the available code lenses through these modules. These are expressed in more detail in the next section since the Wrangler LS has corresponding modules. \label{src:behaviours} 

My task here was 

\begin{itemize}
    \item to initialize and start the Wrangler LS if it enabled in the configuration file;
    \item and for every language feature Wrangler LS provides,
    \begin{itemize}
        \item modify the existing providers to return not just only the responses of the Erlang LS but the Wrangler ones also
        \item or develop a new provider that returns the responses of the Wrangler LS.
    \end{itemize}
\end{itemize}

Namely, I initialized Wrangler in the \tt els\_config:do\_initialize/4\rm\ function, modified the \tt  els\_code\_action\_provider, els\_code\_lens\_provider, els\_document\_highlight\_provider, els\_execute\_command\_provider\rm\ providers, and created and registered a new \tt els\_semantic\_provider\rm\ module. To separate the Wrangler related functions, I made a \tt wrangler\_handler\rm\ utility module.

These tasks are expounded below.

\subsubsection{Initializing Wrangler}

If Wrangler is enabled in the configuration file, it is loaded into the application with the \tt application:load(wrangler)\rm\ call. If Wrangler is not installed on the system, it is first added with the \tt code:add\_path(Path)\rm\ based on the path provided in the configuration file. Lastly, the Wrangler application is started (\tt api\_wrangler:start(\#{wls\_server => true})\rm ). Wrangler's configuration is saved for later use.

The full code snippet complemented with logging is as follows \ref{src:wrangler-init}.

\lstset{caption={Initializing Wrangler in the els\_config:do\_initialize/4 function}, label=src:wrangler-init}
\begin{lstlisting}[language=erlang]
  case maps:get("wrangler", Config, notconfigured) of
    notconfigured -> ok = set(wrangler, notconfigured);
    Wrangler ->
      ok = set(wrangler, Wrangler),
      case maps:get("path", Wrangler, notconfigured) of
        notconfigured -> ?LOG_INFO("Wrangler path is not configured, assuming it is installed system-wide.");
        Path ->
          case code:add_path(Path) of
            true -> ok;
            {error, bad_directory} ->
              ?LOG_INFO("Wrangler path is configured but not a valid ebin directory: ~p", [Path])
          end
      end,
      case application:load(wrangler) of
        ok ->
          case api_wrangler:start(#{wls_server => true}) of
            ok -> ?LOG_INFO("Wrangler started successfully");
            {error, Reason} -> ?LOG_INFO("Wrangler could not be started: ~p", [Reason])
          end;
        {error, Reason} -> ?LOG_INFO("Wrangler could not be loaded: ~p", [Reason])
      end
  end,

\end{lstlisting}


\subsubsection{Modifying existing providers}

Modifying existing providers are presented through the example of code actions.

If Wrangler is enabled, a function defined in Wrangler called \ref{src:code-action-getter} to return the code actions calculated by Wrangler LS. Then, this is appended to the response \ref{src:code-action-provider}. 

If Wrangler is not enabled, neutral elements are appended to the response (an empty list, in particular).

\lstset{caption={Wrangler Code Action getters in the wrangler\_handler module.}, label=src:code-action-getter}
\begin{lstlisting}[language=erlang]
-spec get_code_actions(uri(), range()) -> [map()].
get_code_actions(Uri, Range) ->
  case is_enabled() of % Wrangler is enabled
    true ->
      case wls_code_actions:get_actions(Uri, Range) of % defined in Wrangler
        [] -> [];
        Actions ->
          ?LOG_INFO("Wrangler Code Actions: ~p", [Actions]),
          Actions
      end;
    false -> []
  end.
\end{lstlisting}

\lstset{caption={Appending Wrangler code actions to the original ones in the els\_code\_action\_provider module.}, label=src:code-action-provider}
\begin{lstlisting}[language=erlang]
code_actions(Uri, Range, #{<<"diagnostics">> := Diagnostics}) ->
  lists:flatten([make_code_action(Uri, D) || D <- Diagnostics])
    ++ wrangler_handler:get_code_actions(Uri, Range).
\end{lstlisting}

The other providers are working similarly, although the implementation depends on the response's data types.

\begin{note}
For more transparent Wrangler Forms \todo{missing reference}, in the case of highlights, the response of Wrangler takes precedence over the original response. 
\end{note}

\subsubsection{Creating new providers}

These providers alone are simple as they only return the responses from our \tt wrangler\_handler\rm\ similarly as above \ref{src:code-action-provider}.

The whole provider looks like the following.
\todo{Semantic\_token\_provider code}

Besides from creating a provider that handles the LSP requests, the provider and the new supported language feature has to be registered at the following places: 

\begin{itemize}
    \item \tt els\_methods\rm
    \item \tt els\_general\_provider\rm \todo{what to do there}
    \item \tt els\_provider\rm : the module's name needs to be added in \tt available\_providers/0\rm . 
\end{itemize}


\subsection{The Wrangler Language Server}

As mentioned, Wrangler is loaded into the Erlang Language Server which calls its functions to append the results to the original responses. 

\begin{note}
As Wrangler LS is embedded into Erlang LS, it is not a separate server running apart, contrary to its name. However, Wrangler has its own servers, even one for the Wrangler Forms discussed later \ref{src:wrangler-forms}.
\end{note}

For every refactoring, a module implementing \tt wls\_code\_actions\rm\ has been defined.

An exception to this is the \textit{comment out spec} refactoring as it is implementing the \tt wls\_code\_lens\rm\ behaviour, thus it can be initiated with a code lens instead of code actions. 


In the case of code actions, the full process looks like the following:
\begin{enumerate}
    \item For a given range (set by the cursor position/highlight in the text editor), \tt wrangler\_handler\rm collects the available code actions. For each action returned by \tt wls\_code\_actions:enabled\_actions/0\rm , the \tt wls\_code\_actions:actions/3\rm constructor is called.
    \item The constructor checks the precondition of the given refactoring \todo{something more?}, initializes it, and the returns the code action. This has been done through the code action callbacks presented below \ref{src:code-action-callbacks}.
    \item When the user selects a refactoring, an executeCommand request is generated. The request is forwarded to each other through the \tt els\_execute\_command\_provider, wrangler\_handler, wls\_execute\_command\_provider, wls\_code\_actions \rm modules until it gets handled in the corresponing code action module's \tt execute\_command/1\rm function. This function is also part of the code action callback functions \ref{src:code-action-callbacks}.
    \item Adding additional input is handled by the middleware, hidden in this process line. The middleware's logic is described in the next section \ref{src:middleware}.
    \item Executing the refactorings consists of transforming the given data to Wrangler's refactoring functions' arguments and calling these functions. On failure, error messages are returned to the user with showMessage requests. 
    
\end{enumerate}

The callback modules for the code actions look like the following \ref{src:code-action-callbacks}. As mentioned, the structure and the behaviours is similar to the ones defined in the Erlang Language Server. The differences are mainly caused by the generalization of the utilies that does not apply for refactorings.

\lstset{caption={Callback functions of the wls\_code\_actions behaviour.}, label=src:code-action-callbacks}
\begin{lstlisting}[language=erlang]
%% Title which is shown to the user.
-callback title() -> binary().

%% Identifier of the action and the commands. 
%% Same as the file's name without the `wls_code_action_` prefix.
%% Use `-` instead of `_` in the identifier.
-callback id() -> action_id().

%% Initialize the action. 
%% Optional callback.
-callback init(els_core:uri(), els_core:range()) -> state().

%% Whether the action is offered based on the highlighted range. 
%% Optional callback, defaults to true.
-callback precondition(els_core:uri(), els_core:range()) -> boolean().

%% The command`s arguments.
-callback command_args(els_core:uri(), els_core:range(), state()) -> map().

%% Execute the command with the given arguments. 
%% The first element of the argument list is the one returned by command_args. 
-callback execute_command([any()]) -> [map()].
\end{lstlisting}

These refactorings need to be registered in \tt wls\_code\_actions:available\_actions/0\rm . 
The module names must be the same ones as these values with the \tt wls\_code\_action\_\rm prefix.


\todo{just mention how code lenses work}

\todo{mention wls\_utils to convert between the data representations}

\subsubsection{Wrangler Form}
\label{src:wrangler-forms}

There are some special refactorings for which the solutions discussed above does not work.

These refactorings need more diversified user inputs, in general, the user should be able to see and select one by one the refactoring candidates - the code snippets where the user want to apply the refactoring among all the possibilities.\todo{rephrase}

For these refactorings, we introduced Wrangler Forms: a user interface implemented with LSP features to highlight and select these refactoring candidates.

\todo{description of the cache server, the form's mechanism}

...
Every time a refactoring candidate has been selected, the refactoring takes place, and the form refreshes with one candidate less.
...

\label{src:behaviours}

\subsection{A middleware in the Visual Studio Code extension}
\label{src:middleware}

\todo{description}

\lstset{caption={The command arguments before the middleware}, label=src:middleware-input}
\begin{lstlisting}[language=erlang]
  [#{ 'range' => Range
   , 'uri' => Uri
   , 'user_input' => #{'type' => variable, 'text' => <<"New argument's name">>}
  }].
\end{lstlisting}

\lstset{caption={The command arguments after the middleware}, label=src:middleware-output}
\begin{lstlisting}[language=erlang]
  [#{ 'range' => Range
   , 'uri' => Uri
   , 'user_input' => #{'type' => variable, 'text' => <<"New argument's name">>, 'value' => <<"NewVar">>}
  }].
\end{lstlisting}


\section{Testing}
